---
title: "Getting Started with HiSpaR"
output: 
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting Started with HiSpaR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

High-throughput chromosome conformation capture (Hi-C) has become the gold standard for studying 3D genome organization. While Hi-C generates rich 2D contact maps, a major goal of many genomic studies is to reconstruct the underlying 3D chromatin structures from this data.

HiSpaR is an R package designed to infer these 3D structures using hierarchical Bayesian modeling and Markov Chain Monte Carlo (MCMC) sampling. Unlike traditional constraint-based methods, HiSpaR utilizes a hierarchical framework to efficiently handle large-scale datasets while providing robust uncertainty quantification for the inferred coordinates.

HiSpaR integrates seamlessly with the HiCExperiment ecosystem. Users can easily import Hi-C data from standard formats (such as .mcool and .hic) into HiCExperiment objects and pass them directly to HiSpaR, streamlining the workflow from raw data to 3D visualization.

## Installation

```{r, eval=FALSE}
# Install HiSpaR from Bioconductor
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("HiSpaR")

# HiSpaR depends on several packages for Hi-C data handling
# These are automatically installed with HiSpaR:
# - HiCExperiment: For representing Hi-C data
# - Matrix: For sparse matrix handling

# Additionally, install example data package and HiContacts for data manipulation
BiocManager::install("HiContactsData")
BiocManager::install("HiContacts")
# Optional: Install visualization and analysis packages
install.packages("plotly")  # For interactive 3D visualization
```

### Dependencies

HiSpaR requires the following Bioconductor and CRAN packages:

- **Core**: HiCExperiment, Matrix
- **Suggested**: plotly (for interactive 3D visualization), callr (for safe subprocess execution), HiContacts (for data manipulation), HiContactsData (for example datasets)

## Basic Usage

### Loading Required Packages

```{r setup}
library(HiSpaR)
library(HiContacts)
library(HiContactsData)
library(plotly)

# Load example data from HiCExperiment
cool_file <- CoolFile(HiContactsData::HiContactsData('yeast_wt', format = 'cool'))
hic <- import(cool_file, focus = "II:10000-100000")
# visualize the contact matrix
plotMatrix(hic)
# Load example contact matrix (from HiSpaR's built-in data)
data(su1_contact_mat)
cat("Loaded contact matrix dimensions:", dim(su1_contact_mat), "\n")
```

### Input Data Formats

HiSpaR accepts two types of input:

1. **HiCExperiment objects**: Data from `HiCExperiment::import()` (recommended for Bioconductor integration)
2. **Contact matrices**: Numeric matrices where rows and columns represent genomic loci and entries represent contact counts

Both formats are automatically handled by `hispa_analyze()`, which extracts and processes the contact matrix internally.

### Quick Example

HiSpaR provides the main function `hispa_analyze()` to perform the complete 3D structure inference pipeline:

```{r example, eval=TRUE}
# Run analysis on the example HiCExperiment object
result_yeast <- hispa_analyze(
  hic_experiment = hic,
  output_dir = tempdir(),
  mcmc_iterations = 1000,
  use_cluster_init = TRUE,
  mcmc_burn_in = 10,
  filter_quantile = 0.1,  # Filter out loci in the bottom 10% of contact counts
  verbose = TRUE
)
# Check the structure of results
cat("Result structure:\n")
str(result_yeast)
cat("Final positions (first 5 loci):\n")
print(head(result_yeast$positions, 5))
cat("Estimated parameters: beta0 =", result_yeast$beta0, ", beta1 =", result_yeast$beta1, "\n")

# Run analysis on the example contact matrix
result_su <- hispa_analyze(
  hic_experiment = su1_contact_mat,
  output_dir = tempdir(),
  mcmc_iterations = 2000,
  use_cluster_init = TRUE,
  mcmc_burn_in = 10,
  filter_quantile = 0.1,  # Filter out loci in the bottom 10% of contact counts
  verbose = TRUE
)

# Check the structure of results
cat("Result structure:\n")
str(result_su)
cat("Final positions (first 5 loci):\n")
print(head(result_su$positions, 5))
cat("Estimated parameters: beta0 =", result_su$beta0, ", beta1 =", result_su$beta1, "\n")
```

### Key Parameters

- `mcmc_iterations`: Total number of MCMC iterations (default: 6000)
- `use_cluster_init`: Use hierarchical clustering for initialization (recommended for large datasets)
- `filter_quantile`: Remove loci below this quantile of contact counts (default: -1, no filtering). Use 0.1 to remove bottom 10%
- `mcmc_burn_in`: Number of burn-in iterations to discard (default: 0)

### Visualization

We recommend using the `plotly` package for interactive 3D visualization of the inferred chromatin structures:

```{r visualization, eval=TRUE}
coords_yeast <- result_yeast$positions
# Create interactive 3D scatter plot
plot_ly(
  x = coords_yeast[,1],
  y = coords_yeast[,2],
  z = coords_yeast[,3],
  type = 'scatter3d',
  mode = 'markers+lines',
  marker = list(size = 5, color = ~seq_len(nrow(coords_yeast)), showscale = TRUE),
) %>%
  layout(
    title = "Inferred 3D Chromatin Structure",
    scene = list(
      xaxis = list(title = "X"),
      yaxis = list(title = "Y"),
      zaxis = list(title = "Z")
    )
  )
# Extract coordinates from results
coords_su <- result_su$positions

# Create interactive 3D scatter plot
plot_ly(
  x = coords_su[,1],
  y = coords_su[,2],
  z = coords_su[,3],
  type = 'scatter3d',
  mode = 'markers+lines',
  marker = list(size = 5, color = ~seq_len(nrow(coords_su)), showscale = TRUE),
) %>%
  layout(
    title = "Inferred 3D Chromatin Structure",
    scene = list(
      xaxis = list(title = "X"),
      yaxis = list(title = "Y"),
      zaxis = list(title = "Z")
    )
  )
```

### Output Files

All results are automatically saved to the specified output directory:

- `final_positions.txt`: Final inferred 3D coordinates (n × 3 matrix)
- `initial_positions.txt`: Initial positions before MCMC


## Session Information

```{r sessionInfo}
sessionInfo()
```
